Schema Change: Challenge Problems

### 2024-05-16 paper.tex
Render GL paragraph

Many past improvements in the practice of programming worked by speeding up feedback loops. Live Programming aspires to the optimum of immediate feedback. The problem is that modern software is built out of many layers of different technologies -- thoroughly live programming would propagate changes immediately up and down the entire stack of a running system, from the database to the code, and from the client to the server. We call this \textit{full-stack} live programming. Another problem is that modern software is built with complex collaborations involving development, testing, deployment, and suppport. Such workflows are themselves feedback loops -- throughly live programming would propagate changes between participants frictionlessly. We call this \textit{end-to-end} live programming.

@@

Most programming languages analysis treats state as ephemeral: tied to the lifetime of a running program, and reset when the program is edited. But when we take a broader view, most _programming systems_ (cite) have state that lasts beyond the lifetime of a single execution. This includes collaboration applications with state persisted in a database, image-based systems like Smalltalk, and live programming environments where the program can be edited without resetting all state.

Although these programming systems exist in very different contexts, they all encounter similar problems related to the essential difficulty of preserving state across program edits. New versions of the program must correctly handle pre-existing data. When the data schema or type definitions change, existing data must be evolved to conform to the new schema. Changing data representations or types can also require changing corresponding parts of the program. These problems have been noticed, named, and studied in these different programming systems: "schema evolution" in the databases literature, "hot reloading" in live programming, etc.

>>>>

Many past improvements in the practice of programming worked by speeding up feedback loops. Live Programming aspires to the optimum of immediate feedback. The problem is that modern software is built out of many layers of different technologies -- thoroughly live programming would propagate changes immediately up and down the entire stack of a running system, from the database to the code, and from the client to the server. We call this \textit{full-stack} live programming. Another problem is that modern software is built with complex collaborations involving development, testing, deployment, and suppport. Such workflows are themselves feedback loops -- thoroughly live programming would propagate changes between participants frictionlessly. We call this \textit{end-to-end} live programming.

@@

Most programming languages analysis treats state as ephemeral: tied to the lifetime of a running program, and reset when the program is edited. But when we take a broader view, most \textit{programming systems} (cite) have state that lasts beyond the lifetime of a single execution. This includes collaboration applications with state persisted in a database, image-based systems like Smalltalk, and live programming environments where the program can be edited without resetting all state.

Although these programming systems exist in very different contexts, they all encounter similar problems related to the essential difficulty of preserving state across program edits. New versions of the program must correctly handle pre-existing data. When the data schema or type definitions change, existing data must be evolved to conform to the new schema. Changing data representations or types can also require changing corresponding parts of the program. These problems have been noticed, named, and studied in these different programming systems: ``schema evolution'' in the databases literature, ``hot reloading'' in live programming, etc.
